---
title: "R Summer Assignment -- RMarkdown 3"
author: "YOUR NAME"
output: html_document
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor:
  mode: visual
editor_options:
  markdown:
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

![Kennedy School
Logo](https://www.hks.harvard.edu/sites/default/files/images/paragraph_footer/harvard-kennedy-school-logo_1.png)

**Summer 2024, Pre-Assignment**

**Notebook 3: Introduction to Data Visualization**

The instructions are the same as the first RMarkdown, with a new tutorial link
to watch. Please make a copy of this notebook, watch the tutorial, complete this
document, and submit the URL to your notebook on Canvas.

1.  **Creation**: First, you do not need to make a personal copy of this
    notebook. We encourage you to use the Posit Cloud and, if needed, download
    the project file if you would like to use it offline from your Posit Cloud
    home page.

2.  **Preparation**: Please complete these official RStudio tutorials before
    starting this notebook. We strongly recommend taking notes below and typing
    in the code yourself as you follow this tutorial.

-   **WATCH**: [Visualization Basics](https://vimeo.com/223812632)

This one has several parts, but each part is fairly short. We ask you to
complete them before continuing with the rest of this notebook. There are
several parts this time, so these may take a while. Feel free to complete some
and then come back later.

3.  **Submission**: Complete your assignment in the Posit Cloud environment.
    Please submit only the file you are working on (i.e., your RMarkdown) to
    [this Assignment
    Link](https://www.dropbox.com/request/ciJ01mGyDxrthSSv7cEB). Remember to add
    your name to line 3 of this file.

## Remember: General RMardown Tips

## General Tips

You can enter R code in the code chunks below. To run code, you can click the
play button that appears on the left side of the code chunk when your cursor is
in the box, or press CMD + Enter (or Return) on a Mac, or CTRL + Enter on a PC.

We strongly encourage you to practice writing code as you learn -- do not just
read this document and only complete the exercises. Please write and enter your
own code while you read. Try new things, type code in, and see what happens.
Code as you go. Coding is a skill, much like learning to play an instrument. The
most important part of learning to code is to **practice**!

Remember to save your work frequently. In Posit Cloud, you can save your
notebook by going to File → Save, or with CMD + S (Mac) / Ctrl + S (PC). This
will create a personal copy of the notebook for you to work from.

**Please save often.** While Posit Cloud does autosave, it's a good habit to
manually save your work regularly to ensure you don't lose any progress.

If you get stuck in any aspect of using these Posit Cloud RMarkdowns, please use
Slack to ask the teaching team.

```{r, warning=FALSE}
# INSTALLATION CODE:

# This code block installs a few extra packages you will need
# This may take a few minutes, the icon on the left will spin.
# When it stops spinning it is complete.
# When we get to meet in-person, we are going to learn other forms to load your
# packages.

pkgs <- c("tidyverse", "ggrepel", "gapminder", "maps", "ggthemes")

to_install <- which(!(pkgs %in% rownames(installed.packages())))

install.packages(pkgs[to_install])

library(tidyverse)
library(ggrepel)
library(gapminder)
library(maps)

# Datasets like iris, mpg, gapminder, etc. are all available for you to use here.

# Here, we recreate the Asia dataset in the tutorial for you.
# Filter down to relevant countries
asia <- gapminder |> 
  filter(
    country %in% c("China", "Japan", "Korea, Rep.", "Korea, Dem. Rep."))

# Rename four Asian countries to use
asia <- asia |> 
  mutate(
    country = case_when(
      country == "Korea, Rep." ~ "South Korea",
      country == "Korea, Dem. Rep." ~ "North Korea",
      country == "China" ~ "China",
      country == "Japan" ~ "Japan")
  )

cat("Done!")
```

```{r}
# If needed: 
# TAKE NOTES HERE
```

As usual, once you have completed the RStudio tutorials above, please start here
and continue the document below. These exercises give you additional
opportunities to practice the most important concepts from the RStudio Tutorials
for our HKS courses.

# 1. Building a plot

Looking at a dataset is nice, but we will often want to visualize our data. R
has incredibly powerful tools for data visualization.

To start, load the `tidyverse` library and read this dataset on US presidential
election resulst from 1932 to 2016 by state.

```{r}
# Run this code, loads libraries and does some setup
library(tidyverse)
options(repr.plot.width=10, repr.plot.height=10)
theme_set(theme_gray(base_size = 20))
theme_set(theme_dark(base_size = 20))
theme_set(theme_linedraw(base_size = 30,  ))
update_geom_defaults("point",list(size=5))
update_geom_defaults("line",list(lwd=1.5))

elections <- read_csv("https://www.dropbox.com/s/lhp9nets5qb2rhe/presidential_elections.csv?dl=1")

# show first 10 rows
head(elections, n = 10)
```

Let's look at the results for Massachussetts:

```{r}
# Don't forget to run the code above first to read in the data
ma <- elections |> 
  filter(state == "Massachusetts")

head(ma)

# Practice by creating an object for a different state or time period.
```

As you might expect, there are many functions for plotting! As you learned in
the RStudio tutorials, the starting point for every plot we make in this course
is called `ggplot()`.

Starting with a dataset, you can create a plot with `year` on the x-axis and
`democrat` on the y-axis with:

```{r}
ma |>
  ggplot(
    aes(
      x = year, 
      y = democrat
    )
  )
```

`ggplot()` makes a plot for you, and the `aes()` function (short for
"aesthetic") describes the variables in the dataset that you want on the x and y
axis (for now! We can use `aes()` for other things too later).

But it's empty! To get shapes to appear on the plot, we need to ask for a
particular `geom` (short for "geometry"). A `geom` in R is a way to visualize
the data, like a point, a line, or a shape. To further customize this plot, we
simply add a *geom* for the shape we want. Let's use `geom_line()` to make a
line:

**Hint**: if the plot below looks too small on your computer, you can right
click and "open" it in a new tab to look at. When you start Math Camp, we will
install your own copy of R on your computer and you will be able to create plots
of any size.

```{r}
ma |>
  ggplot(
    aes(
      x = year, 
      y = democrat
    )
  ) +
  geom_line()

# Try creating a line plot for a different state
```

Notice the `+` sign! We add a `+` sign between different pieces of a plot.

We could keep almost this exact code for a plot with a different geometry for
points as well:

```{r}
ma |>
  ggplot(
    aes(
      x = year, 
      y = democrat
    )
  ) +
  geom_point()
```

You can also add both! Notice how the points appear on top of the lines, since
we added them after:

```{r}
ma |>
  ggplot(
    aes(
      x = year, 
      y = democrat
    )
  ) +
  geom_line() +
  geom_point()
```

## Exercises #1

1.  Okay, let's all try this. Create a new object with the election results from
    one state other than Massachusetts. Use it to make a a line plot like we
    have above.

```{r}
# Write your code here 
```

2.  Then, try to make a bar graph using `geom_col()` instead of points or lines.

```{r}
# Write your code here
```

3.  Look back at the line, point, and bar plots you made. Are they all
    displaying the same information? Which one do you think is most effective?

Answer here in text!

# 2. Aesthetics

We added an `x` and `y` aesthetic to choose particular columns to display on our
axes, but plots can accept many other arguments.

### Colors

As you saw in the RStudio tutorials, if you want to make your geoms a certain
color, that is very easy to do with the `color` argument:

```{r}
ma |>
  ggplot(
    aes(
      x = year, 
      y = democrat
    )
  ) +
  geom_line(color = "grey") +
  geom_point(color = "blue")
```

This looks great, but what if we want the colors in our plots to **depend on the
value of the data**? For example, red points for elections that Republicans won
and blue for elections that Democrats won.

Then, people looking at our plot would see additional pieces of information
beyond the values on the x and y axes.

Just like the x and y axes, if we want the color of the points to depend on
values in the data we have to use a **column** in our dataset to define the
colors. Let's make a new column that shows whether the Democratic candidate won
the election.

For a crude measure of the election winner, let's use whether `democrat` is
greater than 50 percent (this is too simple since more than two candidates can
run, but it's okay for now).

```{r}
# Create a new column for a Democratic winner
ma <- ma |>
  mutate(
    winner = democrat > 50
  )

head(ma)
```

Remember how this code works: the column `democrat` in `ma` is really a vector.
The code works very similarly to running something like:

```{r}
democrat <- c(52, 37, 63)
democrat > 50
```

If you want the color of the points to depend on the value of a column, then you
can use the `color` argument in the `aes()` function. R will assign one color to
each value in the `winner` vector. Since there are only `TRUE` and `FALSE`
values in this column, all of the `TRUE` values will have one color and `FALSE`
will have another.

```{r}
ma |>
  ggplot(
    aes(
      x = year, 
      y = democrat, 
      color = winner
    )
  ) +
  geom_point()
```

What if we add the line back?

```{r}
ma |>
  ggplot(
    aes(
      x = year, 
      y = democrat, 
      color = winner)
  ) +
  geom_point() +
  geom_line()
```

Uh-oh! What's happening here? Well, we've asked the plot to change the `color`
of our shapes according to the `winner` variable. Since we have both points and
a line, the plot is trying to change the color both.

What if we only want to change the color of the points depending on the value of
`winner`? Well, we can include that aesthetic **only** in the `geom_point()`
function.

```{r}
ma |>
  ggplot(
    aes(
      x = year, 
      y = democrat
    )
  ) +
  geom_line() +
  geom_point(aes(color = winner))
```

Like before, you can still set the color of the line manually since you don't
want the color to vary by the value of a column. Make sure to do this outside of
`aes()`:

```{r}
ma|>
  ggplot(
    aes(
      x = year, 
      y = democrat
    )
  ) +
  geom_line(color = "grey") +
  geom_point(aes(color = winner))
```

## Size and shape

Similarly, you can have the **size** of a point depend on the value of a column.
For example, see how values with a `winner` value of `TRUE` are larger below
than values with `FALSE`:

```{r}
ma|>
  ggplot(aes(x = year, y = democrat)) +
    geom_line(color = "grey") +
    geom_point(aes(size = winner))
```

Now, points are larger for larger values of `winner`! However, larger values of
`winner` are already higher up on the y-axis, so this does not add much
information to our plot.

The same is true for `shape`:

```{r}
ma|>
  ggplot(
    aes(
      x = year, 
      y = democrat
    )
  ) +
  geom_line(color = "grey") +
  geom_point(aes(shape = winner))
```

## Exercises #2

1.  Create a new column in the `ma` dataset called `percent`. The values should
    be equal the values in `democrat` divided by 100.

```{r}
# Write your code here.
```

2.  Make a scatterplot for the `ma` object with `year` on the x-axis and
    `percent` on the y-axis.

```{r}
# Write your code here.
```

3.  Create a new column in `ma` called `modern` which is `TRUE` for all
    elections after 1980 and `FALSE` for those before. Create a plot with `year`
    on the x-axis, `democrat` on the y-axis, color the points by `winner`, and
    vary the shape by `modern`.

```{r}
# Write your code here.
```

# 3. Customizing your visualizations

Geometries and aesthetics are the core of a nice visualization. R gives you many
many more tools to customize your plots any way you want. For example:

## Labels

**Labels** are important in any plot. We create these with the `labs()`
function, which has arguments for `title`, `subitle`, `caption`, `x`, and `y`
labels. You can choose which labels to include in your plot. For example:

```{r}
ma |>
  ggplot(
    aes(
      x = year, 
      y = democrat
    )
  ) +
  geom_line(color = "grey") +
  geom_point(aes(color = winner)) +
  labs(
    title = "Massachussets Presidential \n Election Results",
    subtitle = "1932-2016",
    x = "Election Year",
    y = "Democratic %"
  )
```

You can also set your own **axes** in R -- the minimum and maximum values on the
x (horizontal) axis and y (vertical) axis. R will often try to pick them for you
automatically, but sometimes you may want to choose your own.

The `xlim()` and `ylim()` functions will take a vector (specificied by `c()`)
with the smallest and largest values you want for that axis.

For example, R automatically chose a y-axis for the previoius plot that
stretched from around 40 to 70 because that's where our values were. However,
what if we wanted to make that go from 0 to 100? We could change `ylim()` like
this:

```{r}
ma |>
  ggplot(
    aes(
      x = year, 
      y = democrat
    )
  ) +
  geom_line(color = "grey") +
  geom_point(aes(color = winner)) +
  labs(title = "Massachussets Presidential \n Election Results",
       subtitle = "1932-2016",
       x = "Election Year",
       y = "Democratic %") +
  ylim(c(0, 100)) # now 0 is the minimum, 100 is the maximum
```

## Themes

**Themes** are simple ways to improve the presentation of your plot as well. We
will learn how to make our own later, but for now you can use built-in themes.
Some built-in themes include `theme_bw()`, `theme_minimal()`, and
`theme_dark()`.

For convenience, you can also store plots to an object and add additional
features onto that object:

```{r}
# save plot in an object called p
p <- ma |>
  ggplot(
    aes(
      x = year, 
      y = democrat
    )
  ) +
  geom_line(color = "grey") +
  geom_point(aes(color = winner)) +
  labs(
    title = "Massachussets Presidential Election Results",
    subtitle = "1932-2016",
    x = "Election Year",
    y = "Democratic %"
  )

# now we can make more customizations to p
# without retyping everything
p + theme_minimal()
p + theme_dark()
```

There are many, many more themes available via packages like `ggthemes`.

```{r}
library(ggthemes)
```

This opens up many many more themes for you, many of which are lisited at [this
link](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/).
Here are a few:

```{r}
p + theme_clean()
p + theme_fivethirtyeight() # 538
p + theme_igray()           # Gray background
p + theme_economist()       # The Economist
p + theme_stata()           # theme from a language called Stata
p + theme_solarized()
```

You can edit almost anything you want about a plot's **theme**, even if you've
already added a preset theme. Most of this works happens through the `theme()`
function. You can run `?theme` to get a full list of options. For example:

```{r}
p +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Facets

Often, you will want to plot several groups at once. However, putting all
information on one plot can be overwhelming. For example, consider this plot:

```{r}
northeast <- elections |> 
  filter(region == "Northeast")

northeast |>
  mutate(winner = democrat > 50) |>
  ggplot(
    aes(
      x = year, 
      y = democrat, 
      color = winner
    )
  ) +
  geom_point()
```

Why is this so cluttered? Well, we are now plotting results from all states in
the Northeast! We could `color` by state instead, but that might look
overwhelming:

```{r}
northeast |>
  ggplot(
    aes(
      x = year, 
      y = democrat,
      color = state
    )
  ) +
  geom_point()
```

Wow! That looks terrible. Instead, what if we plotted a separate line for each
state?

```{r}
northeast |>
  ggplot(
    aes(
      x = year, 
      y = democrat,
      color = state
    )
  ) +
  geom_point() +
  geom_line()
```

That looks a little better, but it is still difficult to tell each line apart
from one another. What if we made a smaller plot for each state and combined
them? This is what a **facet** is. If we ask for a `facet_wrap()` by state, R
will make one plot per state:

```{r}
northeast |>
  ggplot(aes(x = year, y = democrat)) +
  geom_point() +
  geom_line() +
  facet_wrap(~state) + # notice the ~ key (called a tilde)
  theme_linedraw()
```

We could also add the `winner` color back and `facet_wrap()` will automatically
apply it to each plot:

```{r}
northeast |>
  mutate(winner = democrat > 50) |>
  ggplot(
    aes(
      x = year, 
      y = democrat
    )
  ) +
  geom_line() +
  geom_point(aes(color = winner)) +
  facet_wrap(~state) + # notice the ~ key (called a tilde)
  theme_linedraw() +
  labs(
    x = "Election Year",
    y = "Democratic %",
    title = "Presidential Elections",
    subtitle = "1932-2016, Northeastern States"
  )
```

Does the font size look too small to you? There are many many ways of
customizing `ggplot()` objects, many of which we will learn throughout the
course. [Here](https://www.rstudio.com/resources/cheatsheets/) is a helpful
cheatsheet with many of the options listed in case you would like to delve
deeper into this.

```{r}
# For example, functions like axis.text() and axis.title() change font sizes
# for particular places on your plot.
northeast |>
  ggplot(
    aes(
      x = year, 
      y = democrat
    )
  ) +
  geom_point() +
  geom_line() +
  facet_wrap(~state) + # notice the ~ key (called a tilde)
  theme_linedraw() +
  theme(
    strip.text = element_text(size = 25),
    axis.text = element_text(size = 15),
    axis.title = element_text(size = 15)
  )
```

## Exercises #3

1.  The `pop` dataset below contains state population data over time. For any
    state you want, make a plot showing population by year for every year
    **after 1960**.

```{r}
pop <- read_csv("https://www.dropbox.com/s/javbnd4c3n67380/state_population.csv?dl=1")
```

2.  Add labels and a theme to your plot from Question 1.

```{r}
# Write your code here.
```

3.  Now, design a plot (or extend your plot from Question 2) that uses a facet
    in some way (a facet by state or region could be interesting, but feel free
    to be creative!).

```{r}
# Write your code here.
```

## Reminder to Submit

Please follow the submission instructions listed
[here](https://docs.google.com/document/d/195R1jrfbKN5lYZKT2BrU0cJB0Z1JftG8vSZ-60fdi0Q/edit?usp=sharing).
We suggest you submit your assignments as you finish them (i.e., don’t wait
until you have completed them all to submit).

