---
title: "R Summer Assignment -- RMarkdown 1"
output: html_document
date: "2024-05-06"
editor:
  mode: visual
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

![Kennedy School
Logo](https://www.hks.harvard.edu/sites/default/files/images/paragraph_footer/harvard-kennedy-school-logo_1.png)

## Summer 2024, Pre-Assignment

Welcome! If you're here, it means you've already created your GitHub account.
This will be essential for managing your projects and collaborating with others.
For now, we will be using Posit Cloud to complete our summer assignments. You
can also download the entire project to use with RStudio offline, but for the
beginning, we'll focus on the cloud environment.

This set of exercises will help get you up and running with doing basic data
analysis in R. The goal of this assignment is **not** that you will have
perfected R by the time you finish them, but that you have a solid foundation of
key concepts when going into Math Camp. During Math Camp, we will go over much
of this material in greater depth. Spending a significant amount of time on this
now in a way that suits your own schedule will help make your future learning
easier.

RMarkdown is an extension of Markdown used to create dynamic documents that
combine code, output, and text. It allows you to write R code directly in the
document, which can then be executed to display results within the same
document. This integration of code and narrative makes RMarkdown an excellent
tool for creating reproducible reports and interactive documents.

**RMarkdown Notebook 1: Programming Fundamentals**

1.  **Preparation**: Before completing each notebook on this summer assignment,
    we will ask you to please watch a brief primer from the creators of RStudio,
    an application for interacting with R that we will use later in the summer.
    We recommend watching this video first, taking notes in the code box below,
    and coming back later to complete the rest of the exercises.

-   **WATCH**: [Programming Basics](https://rstudio.cloud/learn/primers/1.2)

2.  **Submission**: Complete your assignment in the Posit Cloud environment.
    Please submit only the file you are working on (i.e., your RMarkdown) to the
    Assignment Page on Canvas. We will announce when the Canvas site is created.

## General Tips

You can enter R code in the code chunks below. To run code, you can click the
play button that appears on the left side of the code chunk when your cursor is
in the box, or press CMD + Enter (or Return) on a Mac or CTRL + Enter on a PC.

We strongly encourage you to practice writing code as you learn -- do not just
read this document and only complete the exercises. Please write and enter your
own code while you read. Try new things, type code in, and see what happens.
Code as you go. Coding is a skill, much like learning to play an instrument. The
most important part of learning to code is to **practice**!

In Posit Cloud (and RStudio), you can toggle between Visual and Source modes for
your editor. Visual mode allows for a more WYSIWYG (What You See Is What You
Get) experience, making it easier to format text and visualize results. Source
mode, on the other hand, shows the raw code and markdown, providing greater
control over the document's structure and content. Switching between these modes
can help you understand both the presentation and underlying code of your
notebook.

**Please save often.** Remember to save your work frequently. In Posit Cloud,you
can save your notebook by going to File → Save, or with CMD + S (Mac) / Ctrl + S
(PC).

If you get stuck in any aspect of using these notebooks, please use Slack to ask
the teaching team.

```{r}
# Use this space to write code to take notes on the RStudio video.
sqrt(9)
```

Once you have watched the video above, please continue with the exercises below.
These exercises give you additional opportunities to practice the most important
concepts from the RStudio Tutorials for our HKS courses.

# Data

Throughout this entire course, you will use code to do things with **data**. R
can deal with many types of data, like **numbers** (3, 0, 2021), letters and
words called **characters** ("Data", "South Amboy", "Dom"), and many many
others. For example, in the box below this message, type 3+2 and press the arrow
on the left side (or CMD + Enter on a Mac / Ctrl + Enter on PC):

```{r}
# Try adding below
3+2
```

That was code! R knows that 3 is a number, and it simply reported it back to
you. The number in brackets like this `[5]` simply means that one value was
reported back to you -- R keeps track of these sorts of things for you.

R is also very good at math. Try expanding on the commands written out below:

```{r}
3 + 2
3 - 2
200 * 4
196 / 8

(4 + 2) - (3 * 6)

# Try your own here
```

# Try your own here

You can also type letters and words as long as they are between two quotes
`"like this"`. In the console, type your name and again run the code:

# Type your name below

All of these numbers belong to a certain data type in R called `numeric` while
your name was a `character` (you may have heard terms like class, type, etc. in
reference to certain kinds of data or programming objects, which we will talk
more about later).

Another important type of data in R is **Boolean** values (often called
`logical`), which are `TRUE` or `FALSE`. We often write expressions that will
evaluate to either `TRUE` or `FALSE`. For example, `==` (two equal signs) will
tell you whether or not the value on the left is equal to the value on the
right. If they are equal, the operator returns `TRUE`. If they are not equal,
the operator returns `FALSE`.

# Examples

```{r}
3 == 3 
2 == 4 
5 == 4

# You try it here
```

Now you try it. Is 4 equal to (2 + 2)?

`==` works with characters too -- below, try to see if the character "HKS" equal
to "HBS"

```{r}
# == works on characters too
"Callan" == "Dom"

# You try it here
```

Many other operators in R will evaluate to a Boolean value of `TRUE` or `FALSE`.
For example, `!=` allows you to check if something is NOT equal to something
else. If the values are not equal, R will return `TRUE`. If they are equal, it
will return `FALSE`. Try asking if 4 is not equal to 3.

```{r}
# You try it here
```

## Exercises #1

Below, complete the following exercises using the material above.

1.  Try evaluating a few mathematical expressions in the box below.

```{r}

```

2.  Check that a character for your name is not equal to the character for
    someone else's name.

```{r}

```

3.  Calculate 4017 + 2098. Then, have R confirm that 4017 + 2098 is not equal to
    4016 + 2098.

```{r}

```

4.  **Comments**: you may have noticed the stylish green text in the boxes
    below. These are called "comments," and they are messages that we've left
    for you. Writing your own comments is a very good habit to form, as it makes
    your intentions clear to others, and it will help remind you what your code
    does when you return to it in the future. In R, we use comments by starting
    a line of code with a hashtag: `#`. R will not evaluate these lines as code,
    so they are text that can be inserted into your code.

Go back to #1-3 and add comments above your code.

### Objects

Sometimes, you want to store data for later. An **object** in R will store data
for you. For example, imagine you want to keep the value of a complicated
calculation like the one below:

```{r}
2018 * (6 - 2.5)
```

Instead of trying to remember that number, you can store that data in an
**object**. Then, the data will be there whenever you want it. In R, we make
objects with the `<-` operator. Think of `<-` like an arrow, it stores the
**data on the right** in the **name on the left**. In general, you can give
objects any name you want, as long as there are **no spaces** in the name and it
does not start with a number. For example:

```{r}
result <- 2018 * (6 - 2.5)
```

Wait, nothing happened when you ran that code? This is because your answer is
now stored in an object called `result.` Type `result` in the console and your
answer will come back!

```{r}
result
```

Objects are very powerful tools in R, and can hold any type of data we have
discussed so far. Try making your own below:

```{r}
# Creating objects
name <- "Dom"
name

year <- 2022
year

check_year <- 2022 == (2020 + 2)
check_year
```

In general, you can perform operations on the names of objects the same way you
can perform operations on the value the object contains.

```{r}
# is name equal to Callan?

name == "Callan"

# add 2 to year

year + 2

# is check_year TRUE?

check_year == TRUE
```

## Exercises #2

1.  Create an object named `year` with the value 2021.

```{r}

```

2.  Create an object named `answer` with the value 15. Then, divide `answer` by
    3.  

```{r}

```

3.  Use `==` to see whether `answer` is equal to 12.

```{r}

```

4.  Assign your name to an object called `name`.

```{r}
# prompt: Assign your name to an object called name
```

5.  Try adding 1 to the object containing your name. What happens? Why might
    this be?

```{r}

```

# Functions

Some operations are a little more complex than simple math like addition. A
**function** is a pre-written piece of code that somebody else wrote for you to
perform a specific task. For example, there will be functions for things like
making plots. We will use many functions in this course. We will even learn how
to write our own functions!

For example, how would you find the square root of a number? Well, thankfully
there is a function called `sqrt()`. To run it, you should type `sqrt()` into
the console and put the number that you want to find the square root of between
the parentheses:

```{r}
sqrt(9)

# Take the square root of 16 below
```

This is how all functions will work:

-   Type the name of the function. Here, it is `sqrt()`.
-   Include the value you want to evaluate inside the parentheses (`sqrt(9)`).
    The value you give to the function (`9`) is called an **argument**.

Functions can take **objects** as their argument as well. Remember our object
`number`? Check it out!

```{r}
number <- 16
sqrt(number)
```

You can store the output of a function into an object as well. To see the
result, just type in the name of your object. For example:

```{r}
result <- sqrt(9)
result
```

The `seq()` function will make a sequence of numbers for you. `seq()` takes two
arguments -- the number to start the sequence, and the number to finish it. For
example:

seq(1, 10)

To learn how a function works, you can type `?` followed by the name of the
function (e.g. `?seq`) in the code chunk below to open the **documentation**.

-   **Usage** shows you how to use the function.
-   **Arguments** tells you what arguments you pass the function.
-   **Details** tells you a little bit about the function.
-   **Value** will tell you what the function returns, if anything.
-   **Examples** are very useful, they show you examples of working code.

You may find difficult to parse right now, but with more practice this wil
become second nature!

```{r}
# Open the documentation "help pages" for the seq() function

?seq
```

The `paste()` function is also useful, as it combines two strings:

```{r}
paste("hello", "world")

# Try it below
```

One particularly important function is `ifelse()`. This function is used to
return a different value depending on whether a statement is `TRUE` or `FALSE`.
For example, imagine wanting to change string to represent currency depending on
the country you are working with.

The structure of the function is:

-   `ifelse(condition, value if TRUE, value if FALSE)`

A condition can be anything that evaluates to `TRUE` or `FALSE`. Operators like
`==`, `!=`, `>` (greater than), `<` (less than), `>=` (greater than or equal
to), are very helpful for writing these conditions.

```{r}
# Run this code, then try changing USA to India in the country object, rerun the code, and see what happens.
country <- "USA"
ifelse(country == "USA", "Dollar", "Rupee")
```

```{r}
salary <- 30000
ifelse(salary > 35000, "Higher than 35,000", "Lower than or equal to 35,000")
```

You can also create your own functions. To do so, you need to define a function
just like an object. The syntax for this is a little complicated, but we will
practice it several times throughout the semester.

Study the code below, and then use the function:

```{r}
# Creates a function that adds 10 to a number
  # add_10 is the name off the function
  # x is the "argument"
  # the function adds 10 to x
  # the curly brackets {} take the code inside as the function

add_10 <- function(x) {
  x + 10
}

add_10(4)
add_10(2)
```

Functions can draw on all of the lessons we have learned so far. Here is one
that will be useful for you in a few years:

```{r}
graduate_hks <- function(name, year) {
  paste0(name, ", MPA/ID ", year)
}

graduate_hks("Callan", 2024)

# Try running the function on your name
```

## Exercises #3

1.  Create an object called `number` with the value 1444. What is the square
    root of `number`?

```{r}

```

2.  Now, go in the console and type `?abs` to open the documentation for the
    `abs()` function. Come up with an example that shows what it does.

```{r}

```

3.  Create a function called `is_even()` that returns "Even" if the number is
    even and "Odd" if the number is odd. **Hint**: even numbers have a remainder
    of 0 when divided by 2, while odd numbers do not have a remainder of 0 when
    divided by 2. To find the remainder, use the `%%` operator (this is called
    "modulo"). Call your argument `x` and don't forget to add comments.

```{r}

```

## Example: remainder of 8 divided by 2

```{r}
8 %% 2
```

### Objects with Multiple Values

Objects can also store more than one value at once. **Vectors** are one very
common way to store data. To create a vector, we use the function `c()` (the "c"
stands for combine, because it combines multiple values separated by a comma).
Just like before, you can use the `<-` syntax to create an object.

```{r}
years <- c(2020, 2018, 2007)
names <- c("Callan", "Dom", "Dan")
```

Some functions take **vectors** as arguments because they are meant to work with
multiple values. For example, the `mean()` function will find the average of the
values in a vector:

```{r}
numbers <- c(10, 15)

mean(numbers)
```

We've already learned a lot of terminology. Above, we used a **vector** called
`numbers` as an **argument** in a **function** called `mean()`.

Some operators and functions will run on every entry of the vector separately.
For example:

```{r}
ages <- c(27, 41, 35)
ages + 1
```

## Exercises #4

1.  Create a vector called `prices` with the values 12, 11, and 8. Find the
    average price.

```{r}

```

2.  What happens if you run `prices + 1`? What is this doing? How about
    `prices == 8`?

```{r}

```

3.  The function `sum()` will add all the values in a vector. Run `?sum` to
    learn how it works. Find the total sum of the `prices` object.

```{r}

```

4.  Create a function called `is_sum_greater_than()` that takes two arguments,
    `vector` and `value`. The function should calculate the sum of `vector`,
    check if that sum is greater than `value`, and return "The sum is greater
    than (value)" if it is, and "The sum is NOT greater than (value)" if it is
    not.

```{r}

```

## Reminder to Submit

Please follow the submission instructions listed
[here](https://docs.google.com/document/d/1lt05S6JnBbGTedNLFaLL_b4rmfRTwkW74PK_t7OVuaI/edit).
We suggest you submit notebooks as you finish them (i.e., don’t wait until you
have completed them all to submit).
